Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
reset

---
title: "Reanalysis Step 2: Multiple Imputation"
subtitle: "Imputing missing values using mice with strict temporal and fellow-eye constraints"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

::: {.callout-important}
## Key Change from Previous Version
`cohort_id` has been removed from the imputation model to avoid perfect prediction issues. The variable perfectly predicted outcome missingness (all 'excluded' had missing outcomes, all 'submitted' had observed outcomes), preventing successful imputation. The variable is preserved and added back after imputation for tracking purposes.
:::

@@

# Packages
library(mice)
library(VIM)
library(naniar)
library(dplyr)
library(tidyr)
library(knitr)
library(tibble)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 3 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "test"

if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 3
} else if (analysis_mode == "final") {
  n_imputations <- 30
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}

# Initialize logging
log_file <- init_log("multiple_imputation")
log_message(paste0("Analysis mode: ", analysis_mode))
log_message(paste0("Imputations: ", n_imputations, ", Iterations: ", n_iterations))
```

@@

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars

# Quick record/patient counts
log_message(paste0("Loaded records: ", nrow(df_for_imputation)))
if ("e_mrn_deidentified" %in% names(df_for_imputation)) {
  log_message(paste0("Unique patients: ", dplyr::n_distinct(df_for_imputation$e_mrn_deidentified)))
}
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model (SOURCE variables only)
# Derived variables will be recreated post-imputation

imputation_vars <- c(
  "e_mrn_deidentified",

  # Temporal VA variables
  # NOTE: cohort_id REMOVED - it perfectly predicts outcome missingness
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA
  "outcome_VA_logMAR",         # 2-year outcome
  # NOT outcome_va_vi_binary - derived from logMAR threshold

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"                  # Numeric 0/1/2 → ordered factor
  # NOT dr_severity/person_dr_severity - derived from person_dr
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number(),
    # Store cohort_id separately for later use (not in imputation)
    cohort_id_original = cohort_id
  ) |>
  dplyr::select(row_id, cohort_id_original, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

## Setup Imputation Model (dry run and methods)

```{r setup-mice}
# Dry run to inspect missingness and defaults
mice_dry <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Missingness summary
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  dplyr::filter(N_Missing > 0) |>
  dplyr::arrange(dplyr::desc(N_Missing))

log_message("Variables with missing data:")
print(miss_vars)

# Check that key variables are in the dataset
log_message("\nChecking for key variables in df_mice:")
key_vars_present <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
for (v in key_vars_present) {
  if (v %in% names(df_mice)) {
    n_miss <- sum(is.na(df_mice[[v]]))
    log_message(paste0(v, ": Present (", n_miss, " missing)"))
  } else {
    log_message(paste0(v, ": NOT PRESENT IN DATA"))
  }
}

# Configure methods explicitly
custom_methods <- mice_dry$method

log_message("\nDefault methods from mice:")
print(custom_methods)

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"
# cohort_id removed from dataset to avoid perfect prediction

# Other continuous
custom_methods["CCI"] <- "pmm"
custom_methods["DCSI"] <- "pmm"
custom_methods["age"] <- "pmm"

# Ordinal
custom_methods["person_dr"] <- "polr"

# Binary
custom_methods["catsurg_before_hitplus_cat"] <- "logreg"
custom_methods["ever_lapse_binary"] <- "logreg"  # likely complete, ok to set type
custom_methods["glaucoma_bef_hitplus_cat"] <- "logreg"
custom_methods["otherretina_bef_hitplus_cat"] <- "logreg"
custom_methods["gender_cat"] <- "logreg"  # if binary
custom_methods["other_inject"] <- "logreg"
custom_methods["anti_VEGF"] <- "logreg"
custom_methods["focal_laser_flag"] <- "logreg"
custom_methods["PRP_flag"] <- "logreg"

# Polytomous
custom_methods["race_ethnic_cat"] <- "polyreg"
custom_methods["insurance_cat"] <- "polyreg"

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
custom_methods["cohort_id_original"] <- ""  # Stored for later, not imputed
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

# DIAGNOSTIC: Verify methods are set correctly
log_message("\nMethods for variables with missing data:")
for (v in miss_vars$Variable) {
  if (v %in% names(custom_methods)) {
    log_message(paste0(v, ": ", custom_methods[v]))
  } else {
    log_message(paste0(v, ": NOT IN METHODS VECTOR"))
  }
}

# DIAGNOSTIC: Check which variables will actually be imputed
vars_to_impute_check <- names(custom_methods)[custom_methods != "" & mice_dry$nmis > 0]
log_message(paste0("\nVariables that SHOULD be imputed (method != '' and has missing): ", 
                   length(vars_to_impute_check)))
print(vars_to_impute_check)
```

@@

**Missing data by variable:**
```{r missing-vars-table}
kable(miss_vars)
```

## Setup Predictor Matrix (strict temporal + fellow-eye logic)

```{r setup-predictors}
# Start with empty matrix, then opt in predictors explicitly
pred_matrix <- matrix(0, nrow = ncol(df_mice), ncol = ncol(df_mice))
colnames(pred_matrix) <- names(df_mice)
rownames(pred_matrix) <- names(df_mice)

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified", "cohort_id_original")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

# Never impute but can be used as predictors: ever_lapse_binary
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
# cohort_id removed entirely to avoid perfect prediction issues

# Fellow-eye and helper vars are predictors only (never imputed)
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (fv in fellow_vars) {
  if (fv %in% rownames(pred_matrix)) pred_matrix[fv, ] <- 0
}

# Variable groups
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
global_comorbidity_vars <- c("CCI", "DCSI")
ocular_comorbidity_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", "catsurg_before_hitplus_cat")
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")

# baseline_VA_logMAR predictors (no future info)
if ("baseline_VA_logMAR" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "ever_lapse_binary"] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1
}

# intermediate_VA_logMAR predictors (no treatments, no future)
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix)) {
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "ever_lapse_binary"] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1
  # Explicitly exclude future/fellow-intermediate via omission (remain 0)
}

# outcome_VA_logMAR predictors (allow treatments and fellow baseline/outcome)
if ("outcome_VA_logMAR" %in% rownames(pred_matrix)) {
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "ever_lapse_binary"] <- 1
  for (v in treatment_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
}

# person_dr predictors (pre-outcome only, include fellow_baseline_VA and fellow_dr)
if ("person_dr" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "baseline_VA_logMAR"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["person_dr", "ever_lapse_binary"] <- 1
  for (v in treatment_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_baseline_VA"] <- 1
  if ("fellow_dr" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_dr"] <- 1
}

# Demographics predict each other
for (v1 in demographic_vars) for (v2 in demographic_vars) {
  if (!identical(v1, v2) && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) pred_matrix[v1, v2] <- 1
}

# CCI and DCSI predict each other and are predicted by demographics and exposure
if ("CCI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["CCI", v] <- 1
  if ("DCSI" %in% colnames(pred_matrix)) pred_matrix["CCI", "DCSI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["CCI", "ever_lapse_binary"] <- 1
}
if ("DCSI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["DCSI", v] <- 1
  if ("CCI" %in% colnames(pred_matrix)) pred_matrix["DCSI", "CCI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["DCSI", "ever_lapse_binary"] <- 1
}

# Cataract surgery predicted by demographics/comorbidities/exposure
if ("catsurg_before_hitplus_cat" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", "ever_lapse_binary"] <- 1
}

# Treatment variables predict each other
for (v1 in treatment_vars) for (v2 in treatment_vars) {
  if (!identical(v1, v2) && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) pred_matrix[v1, v2] <- 1
}

log_message("Predictor matrix configured with strict temporal and fellow-eye constraints")

# Save the predictor matrix to CSV for inspection
pred_matrix_df <- as.data.frame(pred_matrix)
pred_matrix_df <- cbind(Variable = rownames(pred_matrix_df), pred_matrix_df)
write.csv(pred_matrix_df, file.path(reanalysis_data_dir, "predictor_matrix.csv"), row.names = FALSE)
log_message(paste0("Saved predictor matrix to: ", file.path(reanalysis_data_dir, "predictor_matrix.csv")))

# Simple verification prints
pred_counts <- data.frame(
  Variable = rownames(pred_matrix),
  N_Predictors = rowSums(pred_matrix > 0),
  N_Times_Used_As_Predictor = colSums(pred_matrix > 0)[rownames(pred_matrix)]
)
pred_counts <- pred_counts[order(-pred_counts$N_Predictors), ]
print(pred_counts)

# Detailed predictors for key variables
for (var in c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")) {
  if (var %in% rownames(pred_matrix)) {
    predictors <- colnames(pred_matrix)[pred_matrix[var, ] == 1]
    log_message(paste0("Predictors for ", var, " (", length(predictors), "): ", paste(predictors, collapse = ", ")))
  }
}
```

@@

```{r final-model-config}
missing_info <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  stringsAsFactors = FALSE
)

final_methods <- data.frame(
  Variable = names(custom_methods),
  Method = custom_methods,
  Type = dplyr::case_when(
    custom_methods == "" ~ "Excluded (not imputed)",
    custom_methods == "pmm" ~ "Predictive mean matching",
    custom_methods == "polr" ~ "Proportional odds regression",
    custom_methods == "logreg" ~ "Logistic regression",
    custom_methods == "polyreg" ~ "Polytomous regression",
    TRUE ~ custom_methods
  ),
  stringsAsFactors = FALSE
) |>
  dplyr::left_join(missing_info, by = "Variable") |>
  dplyr::mutate(
    Will_Be_Imputed = N_Missing > 0 & Method != "",
    Role = dplyr::case_when(
      Variable %in% c("row_id", "e_mrn_deidentified") ~ "ID (excluded)",
      Variable %in% fellow_vars ~ "Fellow (predictor only)",
      N_Missing == 0 & Method != "" ~ "Complete predictor",
      N_Missing > 0 & Method != "" ~ "To be imputed",
      TRUE ~ "Other"
    )
  )

vars_to_impute <- final_methods |>
  dplyr::filter(Will_Be_Imputed) |>
  dplyr::pull(Variable)

kable(final_methods |>
        dplyr::select(Variable, Type, N_Missing, Role) |>
        dplyr::arrange(Role == "ID (excluded)", Role == "Fellow (predictor only)", dplyr::desc(N_Missing)),
      col.names = c("Variable", "Method Type", "Missing Values", "Role in Imputation"))
```

@@

```{r visit-sequence}
visit_sequence <- c()
if ("baseline_VA_logMAR" %in% vars_to_impute) visit_sequence <- c(visit_sequence, "baseline_VA_logMAR")
if ("intermediate_VA_logMAR" %in% vars_to_impute) visit_sequence <- c(visit_sequence, "intermediate_VA_logMAR")
if ("outcome_VA_logMAR" %in% vars_to_impute) visit_sequence <- c(visit_sequence, "outcome_VA_logMAR")
other_vars_to_impute <- setdiff(vars_to_impute, c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR"))
visit_sequence <- c(visit_sequence, other_vars_to_impute)
imputation_order <- paste(visit_sequence, collapse = ", ")
```

@@

```{r imputation-params}
imp_params <- data.frame(
  Parameter = c("Mode", "Number of imputations (m)", "Max iterations (maxit)", "Random seed", "Visit sequence"),
  Value = c(analysis_mode, n_imputations, n_iterations, seed, ifelse(length(visit_sequence) > 0, "Temporal order enforced", "Default"))
)
kable(imp_params)
```

@@

```{r mice-run}
log_message("Starting MICE imputation process...")

# DIAGNOSTIC: Final check before imputation
log_message("\nFinal pre-imputation checks:")
log_message(paste0("Number of rows in df_mice: ", nrow(df_mice)))
log_message(paste0("Number of columns in df_mice: ", ncol(df_mice)))
log_message(paste0("Dimensions of predictor matrix: ", nrow(pred_matrix), " x ", ncol(pred_matrix)))
log_message(paste0("Length of methods vector: ", length(custom_methods)))
log_message(paste0("Length of visit sequence: ", length(visit_sequence)))

# Check alignment
if (ncol(df_mice) != length(custom_methods)) {
  log_message("WARNING: Number of columns doesn't match length of methods vector!")
}
if (ncol(df_mice) != nrow(pred_matrix) || ncol(df_mice) != ncol(pred_matrix)) {
  log_message("WARNING: Predictor matrix dimensions don't match data dimensions!")
}

# Show what will be imputed
log_message("\nVariables in visit sequence:")
print(visit_sequence)

>>>>

---
title: "Reanalysis Step 2: Multiple Imputation"
subtitle: "Imputing missing values using mice package"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Additional packages for imputation
library(mice)
library(VIM)
library(naniar)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 50 imputations, 20 iterations (slow but thorough)
analysis_mode <- "final"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 50
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

**Running in `r toupper(analysis_mode)` mode:**  
- **Imputations:** `r n_imputations`  
- **Iterations:** `r n_iterations`

@@

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model
# Include: outcome, exposure, all covariates source variables only
# Derived variables will be recreated post-imputation

# Core analysis variables - SOURCE ONLY
imputation_vars <- c(
  "e_mrn_deidentified",

  # Temporal VA variables
  "cohort_id",  # Data indicator only - not used as predictor
  "baseline_VA_logMAR",  # Baseline VA
  "intermediate_VA_logMAR",  # Intermediate follow-up VA (for excluded cohort)
  "outcome_VA_logMAR",  # 2-year outcome (true for submitted, missing for excluded)
  # NOT outcome_va_vi_binary - derived from logMAR >= 0.3

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

## Setup Imputation Model

```{r setup-mice}
# Initialize mice
# First, do a dry run to check the imputation setup
mice_dry <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Imputation model summary
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))
```

@@

**Missing data by variable:**
```{r missing-vars-table}
kable(miss_vars)

method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")
```

**Initial imputation methods by variable:**
```{r method-table}
kable(method_table)

# Customize imputation methods as needed
custom_methods <- mice_dry$method

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"  # Impute if missing
custom_methods["intermediate_VA_logMAR"] <- "pmm"  # IMPUTE intermediate values
custom_methods["outcome_VA_logMAR"] <- "pmm"  # Impute 2-year outcomes
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables
custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["age"] <- "pmm"

# Use proportional odds model for ordinal person_dr (0/1/2)
custom_methods["person_dr"] <- "polr"  # Ordinal logistic regression

# Use logistic regression for binary variables
custom_methods["catsurg_before_hitplus_cat"] <- "logreg"

# Use polytomous regression for categorical variables with >2 levels
custom_methods["race_ethnic_cat"] <- "polyreg"
custom_methods["insurance_cat"] <- "polyreg"

# Don't impute these variables
# Method = "" means: don't impute (but still use as predictor EXCEPT for IDs)
custom_methods["row_id"] <- ""  # Row ID - never impute, never use as predictor; see below.
custom_methods["e_mrn_deidentified"] <- ""  # Patient ID - never impute, never use as predictor; see below.

# Variables that are complete (no missing data) - set method but won't actually be imputed
# Setting methods ensures correct measurement level is used IF they had missing data
custom_methods["ever_lapse_binary"] <- "logreg"  # Binary - complete
custom_methods["glaucoma_bef_hitplus_cat"] <- "logreg"  # Binary - already recoded to 0/1
custom_methods["otherretina_bef_hitplus_cat"] <- "logreg"  # Binary - already recoded to 0/1
custom_methods["gender_cat"] <- "logreg"  # Binary categorical - likely complete
custom_methods["other_inject"] <- "logreg"  # Binary - likely complete
custom_methods["anti_VEGF"] <- "logreg"  # Binary - likely complete  
custom_methods["focal_laser_flag"] <- "logreg"  # Binary - likely complete
custom_methods["PRP_flag"] <- "logreg"  # Binary - likely complete

# Note: Variables with complete data won't be imputed even if they have a method assigned
# The method specification ensures proper measurement level if imputation were needed

# Check predictor matrix (which variables predict which)
pred_matrix <- mice_dry$predictorMatrix

# Don't use IDs as predictors AND don't impute them
pred_matrix[, "e_mrn_deidentified"] <- 0
pred_matrix["e_mrn_deidentified", ] <- 0
pred_matrix[, "row_id"] <- 0
pred_matrix["row_id", ] <- 0

# TEMPORAL STRUCTURE RULES
# 1. Baseline can predict intermediate and outcome
pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1

# 2. Intermediate can predict outcome
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

# 3. CRITICAL: Prevent reverse-time prediction
pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict intermediate
pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict baseline
pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0  # Intermediate cannot predict baseline

# Cohort_id should not be predicted by anything AND should not predict anything
pred_matrix["cohort_id", ] <- 0  # Not predicted
pred_matrix[, "cohort_id"] <- 0  # Not a predictor

# Don't use variables as their own predictors
diag(pred_matrix) <- 0

# Ensure exposure is used as predictor for outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

```

@@

```{r final-model-config}

# Get actual missing data information
missing_info <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  stringsAsFactors = FALSE
)

# Summary of methods
final_methods <- data.frame(
  Variable = names(custom_methods),
  Method = custom_methods,
  Type = dplyr::case_when(
    custom_methods == "" ~ "Excluded (IDs - not used as predictor)",
    custom_methods == "pmm" ~ "Predictive mean matching",
    custom_methods == "polr" ~ "Proportional odds regression",
    custom_methods == "logreg" ~ "Logistic regression",
    custom_methods == "polyreg" ~ "Polytomous regression",
    TRUE ~ custom_methods
  ),
  stringsAsFactors = FALSE
) |>
  left_join(missing_info, by = "Variable") |>
  mutate(
    Will_Be_Imputed = N_Missing > 0 & Method != "",
    Role = case_when(
      Variable %in% c("row_id", "e_mrn_deidentified") ~ "ID (excluded)",
      N_Missing == 0 & Method != "" ~ "Complete predictor",
      N_Missing > 0 & Method != "" ~ "To be imputed",
      TRUE ~ "Other"
    )
  )

# Sort to show meaningful groupings
final_methods <- final_methods |>
  dplyr::arrange(Role == "ID (excluded)", N_Missing == 0, Variable)

# Variables that will ACTUALLY be imputed (have missing data AND a method)
vars_to_impute <- final_methods |>
  filter(Will_Be_Imputed) |>
  pull(Variable)
```

**Imputation methods and missing data by variable:**
```{r final-methods-table}
final_methods |>
  select(Variable, Type, N_Missing, Role) |>
  kable(col.names = c("Variable", "Method Type", "Missing Values", "Role in Imputation"))
```

**Variables that will ACTUALLY be imputed** (have missing data, `r length(vars_to_impute)` total):
```{r vars-to-impute}
if (length(vars_to_impute) > 0) {
  data.frame(
    Variable = vars_to_impute,
    Missing_Count = final_methods |> 
      filter(Variable %in% vars_to_impute) |> 
      pull(N_Missing)
  ) |>
  kable()
} else {
  no_imputation_msg <- "No variables require imputation (all data complete)"
  print(no_imputation_msg)
}

# Show complete variables with methods assigned
complete_with_methods <- final_methods |>
  filter(N_Missing == 0 & Method != "") |>
  select(Variable, Type)
```

**Complete variables with imputation methods assigned** (`r nrow(complete_with_methods)` total):
These variables have no missing data and won't be imputed, but have methods specified to ensure proper measurement level:
```{r complete-vars-table}
if (nrow(complete_with_methods) > 0) {
  kable(complete_with_methods, col.names = c("Variable", "Method Type"))
} else {
  no_complete_msg <- "No complete variables with methods assigned"
  print(no_complete_msg)
}

# Show predictor matrix summary
# Get the number of predictors for each variable
predictors_per_var <- rowSums(pred_matrix)

# Separate by variable type
id_vars_pred <- predictors_per_var[c("row_id", "e_mrn_deidentified")]
imputed_vars_pred <- predictors_per_var[vars_to_impute]
other_vars_pred <- predictors_per_var[!names(predictors_per_var) %in% c("row_id", "e_mrn_deidentified", vars_to_impute)]

# Calculate meaningful statistics
n_predictors_for_imputed <- if(length(imputed_vars_pred) > 0) unique(imputed_vars_pred)[1] else NA
predictors <- names(which(colSums(pred_matrix) > 0))

# Verify IDs are excluded
ids_excluded <- all(pred_matrix[, "row_id"] == 0) & all(pred_matrix[, "e_mrn_deidentified"] == 0)
```

### Predictor Matrix Summary

**Number of predictors used:**
- For variables being imputed: **`r if(!is.na(n_predictors_for_imputed)) n_predictors_for_imputed else 0`** predictors each
- For ID variables: **0** predictors (excluded from imputation)

**ID variables properly excluded from predictors:** `r if(ids_excluded) "Yes" else "No"`

**Variables used as predictors** (`r length(predictors)` total):
```{r predictor-list}
if ("row_id" %in% predictors || "e_mrn_deidentified" %in% predictors) {
  warning("WARNING: ID variables are being used as predictors! This should not happen.")
}
print(predictors)
```

@@

```{r visit-sequence}
# Define the order of imputation to respect temporal flow
# Impute earlier timepoints before later ones
visit_sequence <- c()

# First impute baseline if needed
if ("baseline_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "baseline_VA_logMAR")
}

# Then impute intermediate
if ("intermediate_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "intermediate_VA_logMAR")
}

# Then impute outcome
if ("outcome_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "outcome_VA_logMAR")
}

# Add other variables that need imputation
other_vars_to_impute <- vars_to_impute[!vars_to_impute %in% 
                                       c("baseline_VA_logMAR", 
                                         "intermediate_VA_logMAR", 
                                         "outcome_VA_logMAR")]
visit_sequence <- c(visit_sequence, other_vars_to_impute)

# Store for display
imputation_order <- paste(visit_sequence, collapse = ", ")
```

@@

```{r imputation-params}
imp_params <- data.frame(
  Parameter = c("Mode", "Number of imputations (m)", "Max iterations (maxit)", "Random seed", "Visit sequence"),
  Value = c(analysis_mode, n_imputations, n_iterations, seed, "Temporal order enforced")
)
kable(imp_params)
```

@@

```{r run-imputation}
# Run mice imputation
# Note: Parameters are set based on analysis_mode in setup
time_estimate <- if(analysis_mode == "test") "a few minutes" else "15-30 minutes or more"
```

**Starting multiple imputation...**

- Mode: `r analysis_mode`
- Imputations: `r n_imputations`
- Iterations: `r n_iterations`
- Estimated time: `r time_estimate`

```{r mice-run}

mice_results <- mice(
  df_mice,
  m = n_imputations,
  method = custom_methods,
  predictorMatrix = pred_matrix,
  visitSequence = visit_sequence,  # Enforce temporal order
  maxit = n_iterations,  # Uses n_iterations from setup
  seed = 2025,
  printFlag = TRUE
)

# Check convergence
plot(mice_results)
```

## Imputation Diagnostics

**Note:** In the stripplots below, the leftmost column (labeled 1) shows the **original observed data** with missing values. Columns 2 through `r n_imputations + 1` show the imputed datasets.

```{r imputation-diagnostics}
# Check imputed values for key variables
# In mice stripplots: .imp = 1 is original data, .imp = 2+ are imputations

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  # Density plot: Blue = observed data, Red = imputed values
  densityplot(mice_results, ~ CCI, 
              main = "Charlson Comorbidity Index (CCI)") |> print()

  # Stripplot: Column 1 = original data, Columns 2+ = imputed datasets
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
}
